---
title: "Lesson 4 · Sampling, sampling distributions and the central limit theorem"
subtitle: "Part I · Foundations"
description: "Population and sample, the sampling distribution, the standard error, the central limit theorem and the bootstrap."
---

```{r}
#| include: false
source("R/course-data.R")
```

::: {.meta-pills}

[About 2 hours]{} [Before this: Lessons 1–3]{} [Threads: uncertainty]{}

:::

::: {.objectives}

By the end of this lesson you will be able to

- distinguish a population parameter, an estimator and an estimate;
- explain what a sampling distribution is, and what the standard error measures;
- say how the standard error depends on sample size, and why it differs from the standard deviation;
- state what the central limit theorem guarantees, and when it is not enough;
- estimate a standard error by bootstrapping a single sample.

:::

::: {.why}

**Why it matters.** Every study observes one sample out of many it could have drawn. A different sample would give a different estimate. Knowing how much estimates vary from sample to sample is what lets us say how much to trust the one we have. That is the foundation of every confidence interval and p-value in the rest of the course.

:::

## 4.1 Population, sample and estimate {#s1}

Treat the 1,827 days of the city's emergency department record as the **population**. Its mean, the true average number of visits per day, is the **parameter** we want, and here we happen to know it. A study can only afford to look at some days: a **sample**. The sample mean is an **estimator**; its value in one sample is an **estimate**. Draw samples and watch the estimates scatter around the truth.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="sN">Days per sample, n</label><input type="range" id="sN" min="5" max="200" value="20"><output id="sNO"></output></div>
  <div class="btns"><button type="button" id="sOne">Draw 1 sample</button><button type="button" id="sHund">Draw 100 samples</button><button type="button" id="sThou">Draw 1,000 samples</button><button type="button" id="sReset">Start over</button></div>
  <svg id="sdist" viewBox="0 0 520 270" role="img" aria-label="Sampling distribution of the sample mean"></svg>
  <div class="legend"><span><i style="background:var(--p1)"></i>sample means so far</span><span><i style="background:var(--p2)"></i>true population mean</span><span><i style="background:var(--c-ink)"></i>the latest sample</span></div>
</div>
<div class="cpanel"><div class="statgrid" id="sStats"></div><div class="now" id="sNow"></div></div>
</div>
```

::: {.further}

**Further reading:** @kirkwood2003essential on populations, samples and sampling variation; @casella2002statistical on estimators and their properties; @lash2021modern on random and systematic error in epidemiology.

:::

## 4.2 The standard error {#s2}

The spread of the sampling distribution is the **standard error** (SE). It says how far a typical estimate lies from the truth. For a mean, it equals the population SD divided by √n: to halve the SE you need four times as many observations. The curve shows the SE for every sample size; the dot is where your current n sits, and the cross is the spread actually observed among your simulated samples.

```{=html}
<div class="cw grid2">
<div class="cpanel"><svg id="sePlot" viewBox="0 0 520 260" role="img" aria-label="Standard error against sample size"></svg></div>
<div class="cpanel"><div class="now" id="seNow"></div></div>
</div>
```

::::: {.triad}

:::: {.a}

::: {.lab}

[A]{.k}Standard notation

:::

$$SE(\bar{X}) = \frac{\sigma}{\sqrt{n}}, \qquad \widehat{SE} = \frac{s}{\sqrt{n}}$$

::: {.note}

$\sigma$ = population SD, $s$ = sample SD, $n$ = sample size.

:::

::::

:::: {.b}

::: {.lab}

[B]{.k}Model list and R

:::

```r
smp <- city_daily |> slice_sample(n = 20)
smp |>
  summarise(mean = mean(visits),
            sd   = sd(visits),
            se   = sd / sqrt(n()))
```

::::

:::: {.c}

::: {.lab}

[C]{.k}How it's reported

:::

[ ]{#seReport}

::: {.note}

Report the SD to describe the data and the SE (or better, a confidence interval) to describe the estimate.

:::

::::

:::::

::: {.callout-warning title="Common pitfall"}

Reporting "mean ± SE" to describe variability in the data. The SE shrinks as the sample grows, so it makes data look less variable than they are. Use the SD to describe the data and the SE for the precision of the mean.

:::

::: {.callout-note collapse="true" appearance="simple" icon="false" title="R code (optional): simulating the sampling distribution"}

```{r}
city_daily <- visits |>
  group_by(date) |>
  summarise(visits = sum(visits), pm25 = first(pm25), .groups = "drop")

set.seed(1)
sample_means <- map_dbl(1:2000, \(i) mean(slice_sample(city_daily, n = 20)$visits))

tibble(true_mean = mean(city_daily$visits),
       mean_of_estimates = mean(sample_means),
       empirical_se = sd(sample_means),
       theoretical_se = sd(city_daily$visits) / sqrt(20))
```

:::

::: {.further}

**Further reading:** @altman2005standard explain the difference between standard deviations and standard errors; @kirkwood2003essential derive the standard error of a mean and of a proportion.

:::

## 4.3 The central limit theorem {#s3}

Daily PM2.5 is right-skewed. Yet if you average enough days, the averages follow a nearly **normal** distribution: this is the **central limit theorem**. It is why so many intervals use 1.96. How many days are "enough" depends on how skewed the data are. Change the number of days averaged.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="cVar">Variable</label><select id="cVar"><option value="pm">daily PM2.5 (right-skewed)</option><option value="lake">daily visits, Lakeside 0–14 (small counts)</option><option value="temp">daily temperature (two peaks)</option></select></div>
  <div class="btns" id="cN" role="group" aria-label="Values averaged"><button type="button" data-n="1" aria-pressed="false">n = 1</button><button type="button" data-n="2" aria-pressed="false">2</button><button type="button" data-n="5" aria-pressed="true">5</button><button type="button" data-n="10" aria-pressed="false">10</button><button type="button" data-n="30" aria-pressed="false">30</button><button type="button" data-n="100" aria-pressed="false">100</button></div>
  <svg id="clt" viewBox="0 0 520 260" role="img" aria-label="Distribution of averages with a normal curve"></svg>
  <div class="legend"><span><i style="background:var(--c-lik)"></i>2,000 averages of n days</span><span><i style="background:var(--p1)"></i>normal curve with the same mean and SD</span></div>
</div>
<div class="cpanel"><div class="statgrid" id="cStats"></div><div class="now" id="cNow"></div></div>
</div>
```

::: {.further}

**Further reading:** @casella2002statistical state and prove the central limit theorem; @blitzstein2019introduction give an accessible treatment; @kirkwood2003essential discuss when the normal approximation is adequate.

:::

## 4.4 The bootstrap: a standard error from one sample {#s4}

In a real study there is only one sample. The **bootstrap** imitates repeated sampling by resampling from the sample itself, with replacement, many times, and recomputing the estimate each time. The spread of these bootstrap estimates approximates the standard error. Here, one sample of 30 children estimates mean residential NO₂.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="bN">Children in the sample</label><input type="range" id="bN" min="10" max="300" value="30"><output id="bNO"></output></div>
  <div class="btns"><button type="button" id="bNew">Draw a new study sample</button><button type="button" id="bRun">Run 2,000 bootstrap resamples</button></div>
  <svg id="boot" viewBox="0 0 520 260" role="img" aria-label="Bootstrap distribution of the mean"></svg>
  <div class="legend"><span><i style="background:var(--p4)"></i>bootstrap means</span><span><i style="background:var(--c-ink)"></i>the sample's own mean</span><span><i style="background:var(--p2)"></i>true mean in all 5,000 children</span></div>
</div>
<div class="cpanel"><div class="statgrid" id="bStats"></div><div class="now" id="bNow"></div></div>
</div>
```

::::: {.triad}

:::: {.a}

::: {.lab}

[A]{.k}Standard notation

:::

$$y^{*(b)} \sim \text{resample}(y_1, \dots, y_n), \quad b = 1, \dots, B$$
$$\widehat{SE}_{\text{boot}} = \mathrm{SD}\big(\hat\theta^{*(1)}, \dots, \hat\theta^{*(B)}\big)$$

::::

:::: {.b}

::: {.lab}

[B]{.k}Model list and R

:::

```r
smp <- cohort |> slice_sample(n = 30)
boot <- map_dbl(1:2000, \(b)
  mean(sample(smp$no2_modeled, replace = TRUE)))
sd(boot)                         # bootstrap SE
quantile(boot, c(0.025, 0.975))  # percentile CI
```

::::

:::: {.c}

::: {.lab}

[C]{.k}How it's reported

:::

[ ]{#bReport}

::::

:::::

::: {.callout-note collapse="true" appearance="simple" icon="false" title="R code (optional): a bootstrap standard error"}

```{r}
set.seed(2)
smp  <- cohort |> slice_sample(n = 30)
boot <- map_dbl(1:2000, \(b) mean(sample(smp$no2_modeled, replace = TRUE)))

tibble(estimate     = mean(smp$no2_modeled),
       se_formula   = sd(smp$no2_modeled) / sqrt(30),
       se_bootstrap = sd(boot),
       ci_low       = quantile(boot, 0.025),
       ci_high      = quantile(boot, 0.975))
```

:::

::: {.callout-tip icon="false" title="Bayesian preview"}

The sampling distribution is a frequentist idea: it describes what would happen across studies you never ran. A Bayesian instead describes uncertainty about the parameter given the one sample you have, with a posterior distribution. With little prior information and enough data, the two give similar intervals; Lesson 19 shows when and why they differ.

:::

::: {.further}

**Further reading:** @efron1979bootstrap introduced the bootstrap; @efron1993introduction is the standard book; @carpenter2000bootstrap is a practical guide to bootstrap confidence intervals for medical statisticians.

:::

## Check your understanding {#quiz}

```{=html}
<div class="cw" id="quizzes"></div>
```

## Key takeaways {#take}

- A parameter is fixed; an estimate varies from sample to sample. The sampling distribution describes that variation.
- The standard error is the SD of the sampling distribution. For a mean, SE = σ/√n.
- The SD describes the data; the SE describes the precision of an estimate. They answer different questions.
- Averages of many values are approximately normal (central limit theorem), sooner for symmetric data than for skewed data.
- The bootstrap estimates a standard error from a single sample by resampling it.

```{=html}
<div class="done" data-lesson="lesson-04"><button type="button"></button><span></span></div>
<script src="assets/lesson-04.js"></script>
```
