---
title: "Lesson 8 · Generalized linear models: logistic and Poisson regression"
subtitle: "Part II · Frequentist regression"
description: "Link functions, logistic regression and odds ratios, Poisson regression with offsets, and overdispersion."
---

```{r}
#| include: false
source("R/course-data.R")
```

::: {.meta-pills}

[About 3 hours]{} [Before this: Lessons 3, 5 and 7]{} [Threads: likelihood · uncertainty]{}

:::

::: {.objectives}

By the end of this lesson you will be able to

- describe a GLM as a distribution, a link function and a linear predictor;
- fit and interpret a logistic regression, and report odds ratios per a sensible exposure unit;
- estimate risk ratios for a common binary outcome;
- fit a Poisson regression for rates with an offset, and interpret rate ratios;
- detect overdispersion and correct standard errors with a quasi-Poisson model.

:::

::: {.why}

**Why it matters.** Most epidemiological outcomes are binary (asthma yes or no) or counts (visits per day). Linear regression handles neither well. Generalized linear models keep the familiar "outcome = function of predictors" structure, but let the mean be a probability or a rate, and let the variance follow the data type. The odds ratios and rate ratios in most papers come from them.

:::

## 8.1 Three ingredients: distribution, link, linear predictor {#s1}

A GLM has a **linear predictor**, η = α + βx, exactly as in linear regression. A **link function** connects η to the mean, so the mean stays in its allowed range: the log link keeps counts positive, the logit link keeps probabilities between 0 and 1. And a **distribution** describes how observations scatter around the mean. Change α and β and see the same linear predictor through three links.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="kA">Intercept α</label><input type="range" id="kA" min="-3" max="3" step="0.1" value="-1.8"><output id="kAO"></output></div>
  <div class="ctrl"><label for="kB">Slope β</label><input type="range" id="kB" min="-1" max="1.5" step="0.05" value="0.5"><output id="kBO"></output></div>
  <svg id="links" viewBox="0 0 520 300" role="img" aria-label="Mean under identity, log and logit links"></svg>
  <div class="legend"><span><i style="background:var(--c-lik)"></i>identity: mean = η</span><span><i style="background:var(--p1)"></i>log: mean = e<sup>η</sup></span><span><i style="background:var(--p2)"></i>logit: mean = 1 / (1 + e<sup>−η</sup>)</span></div>
</div>
<div class="cpanel"><div class="now" id="kNow"></div></div>
</div>
```

::::: {.triad}

:::: {.a}

::: {.lab}

[A]{.k}Standard notation

:::

$$Y_i \sim \text{distribution}(\mu_i), \qquad g(\mu_i) = \eta_i = \alpha + \beta x_i$$

::: {.note}

Linear regression: normal, $g(\mu) = \mu$. Logistic: Bernoulli, $g(\pi) = \log\frac{\pi}{1-\pi}$. Poisson: $g(\mu) = \log\mu$.

:::

::::

:::: {.b}

::: {.lab}

[B]{.k}Model list and R

:::

```r
glm(y ~ x, family = gaussian)              # identity
glm(y ~ x, family = binomial(link = "logit"))
glm(y ~ x, family = poisson(link = "log"))
```

::::

:::: {.c}

::: {.lab}

[C]{.k}How it's reported

:::

Coefficients are reported on the scale that makes sense for the link: as differences for the identity link, **odds ratios** $e^{\beta}$ for the logit link, and **rate ratios** $e^{\beta}$ for the log link.

::::

:::::

::: {.further}

**Further reading:** @mccullagh1989generalized is the original book-length treatment of GLMs; @agresti2013categorical and @fox2016applied give accessible introductions.

:::

## 8.2 Logistic regression and odds ratios {#s2}

Logistic regression models the log-odds of asthma as a linear function of NO₂ and other variables. Its slope, exponentiated, is an **odds ratio**. Because a 1 µg/m³ change is trivial, report the OR per a meaningful unit, such as 10 µg/m³ or an interquartile range. Compare the crude and adjusted estimates, and the estimate using modeled NO₂ with the one using the true exposure, known here only because the data are simulated.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="gU">Report per</label><select id="gU"><option value="10">10 µg/m³</option><option value="1">1 µg/m³</option><option value="iqr">interquartile range</option></select></div>
  <div class="ctrl"><label for="gX">Exposure</label><select id="gX"><option value="mod">modeled NO₂ (what a real study has)</option><option value="true">true NO₂ (simulation only)</option></select></div>
  <label class="small"><input type="checkbox" id="gAdj"> adjust for SES, parental smoking, green space and sex</label>
  <svg id="gPlot" viewBox="0 0 520 280" role="img" aria-label="Observed and fitted asthma risk by NO2"></svg>
  <div class="legend"><span><i style="background:var(--c-ink);border-radius:50%;width:9px"></i>observed risk in tenths of NO₂</span><span><i style="background:var(--p2)"></i>fitted logistic curve</span></div>
</div>
<div class="cpanel"><div class="statgrid" id="gStats"></div><div class="tw"><table data-quarto-disable-processing="true" class="t" id="gTab"></table></div><div class="now" id="gNow"></div></div>
</div>
```

::::: {.triad}

:::: {.a}

::: {.lab}

[A]{.k}Standard notation

:::

$$Y_i \sim \mathrm{Bernoulli}(\pi_i)$$
$$\mathrm{logit}(\pi_i) = \alpha + \beta\,\frac{\mathrm{NO}_{2,i}}{10} + \gamma^\top z_i, \qquad OR_{10} = e^{\beta}$$

::::

:::: {.b}

::: {.lab}

[B]{.k}Model list and R

:::

```r
asthma[i] ~ bernoulli(p[i])
logit(p[i]) = a + b * no2[i] / 10 + g * z[i]

fit <- glm(asthma ~ I(no2_modeled / 10) + ses +
             parent_smoke + green + sex,
           family = binomial, data = cohort)
broom::tidy(fit, exponentiate = TRUE,
            conf.int = TRUE)
```

::::

:::: {.c}

::: {.lab}

[C]{.k}How it's reported

:::

[ ]{#gReport}

::::

:::::

::: {.callout-warning title="Common pitfall"}

Reporting an OR per 1 µg/m³ (for example 1.014) with too few decimals, so it reads as "no effect", or presenting an adjusted OR for a common outcome as if it were a risk ratio. Choose a meaningful unit, and estimate risk ratios directly when the outcome is common.

:::

::: {.callout-note collapse="true" appearance="simple" icon="false" title="R code (optional): logistic regression, and risk ratios by modified Poisson"}

```{r}
fit_crude <- glm(asthma ~ I(no2_modeled / 10), family = binomial, data = cohort)
fit_adj   <- glm(asthma ~ I(no2_modeled / 10) + ses + parent_smoke + green + sex,
                 family = binomial, data = cohort)
bind_rows(crude = broom::tidy(fit_crude, exponentiate = TRUE, conf.int = TRUE)[2, ],
          adjusted = broom::tidy(fit_adj, exponentiate = TRUE, conf.int = TRUE)[2, ],
          .id = "model")

# risk ratio per 10 ug/m3: Poisson model with robust (sandwich) standard errors
fit_rr <- glm(asthma ~ I(no2_modeled / 10) + ses + parent_smoke + green + sex,
              family = poisson, data = cohort)
rr_se <- sqrt(diag(sandwich::vcovHC(fit_rr, type = "HC0")))[2]
exp(coef(fit_rr)[2] + c(estimate = 0, low = -1.96, high = 1.96) * rr_se)
```

:::

::: {.further}

**Further reading:** @hosmer2013applied is the standard text on logistic regression; @zou2004modified introduced modified Poisson regression for risk ratios; @altman1998odds on why odds ratios mislead for common outcomes.

:::

## 8.3 Poisson regression for rates: the offset {#s3}

District rates of emergency visits depend on age structure and deprivation. A Poisson regression of visit counts must account for how many person-years each count comes from. That is the role of the **offset**: log(person-years) enters the model with a coefficient fixed at 1, so the model describes rates, not raw counts. Each row below is one district and age group over five years. Remove the offset and see what goes wrong.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <label class="small"><input type="checkbox" id="oOff" checked> include log(person-years) as an offset</label><br>
  <label class="small"><input type="checkbox" id="oDep" checked> include district deprivation</label>
  <div class="tw"><table data-quarto-disable-processing="true" class="t" id="oTab"></table></div>
</div>
<div class="cpanel"><div class="now" id="oNow"></div></div>
</div>
```

::::: {.triad}

:::: {.a}

::: {.lab}

[A]{.k}Standard notation

:::

$$D_{jk} \sim \mathrm{Poisson}(\mu_{jk})$$
$$\log \mu_{jk} = \log PT_{jk} + \alpha + \gamma_k + \beta\,\mathrm{dep}_j$$

::: {.note}

$j$ = district, $k$ = age group. $e^{\gamma_k}$ and $e^{\beta}$ are rate ratios.

:::

::::

:::: {.b}

::: {.lab}

[B]{.k}Model list and R

:::

```r
D[j,k] ~ poisson(mu[j,k])
log(mu[j,k]) = log(PT[j,k]) + a + g[k] + b * dep[j]

agg <- visits |>
  group_by(district_id, age_group) |>
  summarise(D = sum(visits),
            PT = sum(population) / 365.25) |>
  left_join(districts, by = "district_id")
glm(D ~ age_group + deprivation + offset(log(PT)),
    family = poisson, data = agg)
```

::::

:::: {.c}

::: {.lab}

[C]{.k}How it's reported

:::

[ ]{#oReport}

::::

:::::

::: {.callout-note collapse="true" appearance="simple" icon="false" title="R code (optional): rate ratios from a Poisson model with an offset"}

```{r}
agg <- visits |>
  group_by(district_id, age_group) |>
  summarise(D = sum(visits), PT = sum(population) / 365.25, .groups = "drop") |>
  left_join(districts, by = "district_id") |>
  mutate(age_group = factor(age_group, levels = c("0-14", "15-64", "65+"), ordered = FALSE))

fit_rate <- glm(D ~ age_group + deprivation + offset(log(PT)), family = poisson, data = agg)
broom::tidy(fit_rate, exponentiate = TRUE, conf.int = TRUE)
```

:::

::: {.further}

**Further reading:** @frome1985use on Poisson regression for incidence rates; @clayton1993statistical on rates, person-time and log-linear models; @cameron2013regression on count models.

:::

## 8.4 Overdispersion and quasi-Poisson standard errors {#s4}

A Poisson model assumes the variance equals the mean. Daily emergency counts vary more (Lesson 3). The coefficients are still estimated correctly, but their standard errors are too small. The **quasi-Poisson** model estimates a dispersion parameter φ and multiplies every standard error by √φ. Here the question is the effect of PM2.5 on daily visits, adjusted for season and day of the week.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="qS">Series</label><select id="qS"><option value="city">whole city, all ages</option><option value="ot">Old Town, 65+</option><option value="lk">Lakeside, 0–14</option></select></div>
  <label class="small"><input type="checkbox" id="qSea" checked> adjust for season and day of week</label>
  <svg id="qPlot" viewBox="0 0 520 220" role="img" aria-label="Poisson and quasi-Poisson confidence intervals"></svg>
</div>
<div class="cpanel"><div class="statgrid" id="qStats"></div><div class="now" id="qNow"></div></div>
</div>
```

::::: {.triad}

:::: {.a}

::: {.lab}

[A]{.k}Standard notation

:::

$$E[Y_t] = \mu_t, \qquad \mathrm{Var}(Y_t) = \phi\,\mu_t$$
$$\hat\phi = \frac{1}{n-p}\sum_t \frac{(y_t - \hat\mu_t)^2}{\hat\mu_t}, \qquad SE_{\text{quasi}} = \sqrt{\hat\phi}\,SE_{\text{Poisson}}$$

::::

:::: {.b}

::: {.lab}

[B]{.k}Model list and R

:::

```r
glm(visits ~ I(pm25 / 10) + sin(2*pi*doy/365.25) +
      cos(2*pi*doy/365.25) + dow,
    family = quasipoisson, data = city_daily)
```

::::

:::: {.c}

::: {.lab}

[C]{.k}How it's reported

:::

[ ]{#qReport}

::::

:::::

::: {.callout-tip icon="false" title="Bayesian preview"}

In a Bayesian model, overdispersion is handled by choosing a negative binomial likelihood, whose extra parameter is estimated with everything else, or by adding a random effect for each day. Posterior predictive checks (Lesson 22) reveal overdispersion the model misses.

:::

::: {.further}

**Further reading:** @verhoef2007quasi compare quasi-Poisson with negative binomial models; @mccullagh1989generalized introduced quasi-likelihood for GLMs; @agresti2013categorical discusses overdispersion in count and binary data.

:::

## Check your understanding {#quiz}

```{=html}
<div class="cw" id="quizzes"></div>
```

## Key takeaways {#take}

- A GLM = a distribution + a link function + a linear predictor. Linear, logistic and Poisson regression are all GLMs.
- In logistic regression, $e^{\beta}$ is an odds ratio. Report it per a meaningful unit of exposure.
- For common outcomes, estimate risk ratios directly (for example, Poisson regression with robust standard errors).
- In Poisson regression for rates, log(person-time) is an offset. Without it, a model describes counts, not rates.
- Overdispersion leaves estimates unbiased but makes Poisson standard errors too small; quasi-Poisson or negative binomial models fix this.

```{=html}
<div class="done" data-lesson="lesson-08"><button type="button"></button><span></span></div>
<script src="assets/lesson-08.js"></script>
```
