---
title: "Lesson 11 · Nonlinear effects: categories, polynomials, splines and GAMs"
subtitle: "Part II · Frequentist regression"
description: "Why categorizing a continuous exposure loses information, polynomials, regression splines, and penalized splines in generalized additive models."
---

```{r}
#| include: false
source("R/course-data.R")
```

::: {.meta-pills}

[About 3 hours]{} [Before this: Lessons 8 and 10]{} [Threads: shrinkage]{}

:::

::: {.objectives}

By the end of this lesson you will be able to

- explain what is lost when a continuous exposure is cut into categories;
- describe why polynomials behave badly at the edges of the data;
- build a regression spline from basis functions and choose its flexibility;
- explain a penalized spline as a flexible curve whose wiggliness is restrained by the data;
- present a nonlinear exposure–response function relative to a reference value.

:::

::: {.why}

**Why it matters.** Temperature affects health in a U shape: cold days and hot days both bring more emergency visits. A single slope can't describe that. How you model the curve decides what you conclude about the minimum-risk temperature, the size of heat effects, and the burden attributable to heat (Lessons 14 and 16).

:::

Throughout this lesson the outcome is daily city-wide emergency visits, modeled by quasi-Poisson regression adjusted for season and day of week. The exposure is the average temperature over the same day and the three before it. (Lesson 16 models the lags properly.)

## 11.1 Categories: the step function {#s1}

The simplest way to allow a nonlinear effect is to cut temperature into categories and estimate a rate ratio for each, relative to a reference category. The result is a **step function**: flat within each category, with jumps at arbitrary cut-points. Days just either side of a cut-point are treated as completely different; days at opposite ends of a category as identical.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="cK">Number of categories</label><input type="range" id="cK" min="2" max="12" value="4"><output id="cKO"></output></div>
  <div class="ctrl"><label for="cT">Cut-points at</label><select id="cT"><option value="q">quantiles (equal numbers of days)</option><option value="e">equal-width temperature bands</option></select></div>
  <svg id="cPlot" viewBox="0 0 520 300" role="img" aria-label="Rate ratios by temperature category"></svg>
</div>
<div class="cpanel"><div class="statgrid" id="cStats"></div><div class="now" id="cNow"></div></div>
</div>
```

::: {.further}

**Further reading:** @bennette2012against argue against quantile categories in epidemiology; @royston2006dichotomizing on the costs of dichotomizing; @greenland1995dose on alternatives to categorical dose–response analysis.

:::

## 11.2 Polynomials and their wild tails {#s2}

Adding T², T³ and so on lets the curve bend smoothly. But polynomials are **global**: every coefficient affects the whole curve, so data in the middle of the range dictate behavior at the extremes, where data are sparse. High-degree polynomials swing wildly at the edges, exactly where heat and cold effects matter most.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="pD">Polynomial degree</label><input type="range" id="pD" min="1" max="8" value="2"><output id="pDO"></output></div>
  <svg id="pPlot" viewBox="0 0 520 300" role="img" aria-label="Polynomial exposure-response curve"></svg>
</div>
<div class="cpanel"><div class="statgrid" id="pStats"></div><div class="now" id="pNow"></div></div>
</div>
```

::: {.further}

**Further reading:** @harrell2015regression on why polynomials and categories are inferior to splines; @fox2016applied on polynomial regression; @greenland1995dose on modeling dose–response trends.

:::

## 11.3 Regression splines: local flexibility {#s3}

A **spline** joins polynomial pieces smoothly at points called **knots**. A natural (restricted) cubic spline is also forced to be straight beyond the outer knots, which tames the tails. It enters the model as a set of **basis functions**: new columns in the design matrix, each multiplied by its own coefficient. The fitted curve is their weighted sum. More knots mean more flexibility.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="sK">Knots</label><input type="range" id="sK" min="3" max="7" value="4"><output id="sKO"></output></div>
  <svg id="sPlot" viewBox="0 0 520 300" role="img" aria-label="Spline exposure-response curve"></svg>
  <svg id="sBasis" viewBox="0 0 520 170" role="img" aria-label="Spline basis functions"></svg>
</div>
<div class="cpanel"><div class="statgrid" id="sStats"></div><div class="now" id="sNow"></div></div>
</div>
```

::::: {.triad}

:::: {.a}

::: {.lab}

[A]{.k}Standard notation

:::

$$\log \mu_t = \alpha + f(T_t) + \gamma^\top z_t, \qquad f(T) = \sum_{k=1}^{K-1} \beta_k B_k(T)$$
$$RR(T) = \exp\{f(T) - f(T_{\text{ref}})\}$$

::: {.note}

$B_k$ are the spline basis functions; $z_t$ the season and day-of-week terms; $T_{\text{ref}}$ the reference temperature.

:::

::::

:::: {.b}

::: {.lab}

[B]{.k}Model list and R

:::

```r
visits[t] ~ quasipoisson(mu[t])
log(mu[t]) = a + f(temp[t]) + season + dow

fit <- glm(visits ~ splines::ns(temp03, df = 3) +
             sin(2*pi*doy/365.25) +
             cos(2*pi*doy/365.25) + dow,
           family = quasipoisson, data = city_daily)
```

::::

:::: {.c}

::: {.lab}

[C]{.k}How it's reported

:::

[ ]{#sReport}

::::

:::::

::: {.callout-note collapse="true" appearance="simple" icon="false" title="R code (optional): a spline exposure–response curve relative to its minimum"}

```{r}
city_daily <- visits |>
  group_by(date) |>
  summarise(visits = sum(visits), temp = first(temp_city), dow = first(dow), .groups = "drop") |>
  mutate(doy = yday(date),
         temp03 = (temp + lag(temp, 1) + lag(temp, 2) + lag(temp, 3)) / 4) |>
  filter(!is.na(temp03))

fit <- glm(visits ~ splines::ns(temp03, df = 3) + sin(2*pi*doy/365.25) + cos(2*pi*doy/365.25) + dow,
           family = quasipoisson, data = city_daily)

grid <- tibble(temp03 = seq(-5, 30, by = 0.5), doy = 180, dow = factor("Wed", levels = levels(city_daily$dow)))
grid$logmu <- predict(fit, newdata = grid)
grid |> mutate(rr = exp(logmu - min(logmu))) |>
  ggplot(aes(temp03, rr)) + geom_line() + geom_hline(yintercept = 1, linetype = 2) +
  labs(x = "Mean temperature, lags 0-3 (°C)", y = "Rate ratio vs minimum")
```

:::

::: {.further}

**Further reading:** @perperoglou2019review compare spline procedures in R; @harrell2015regression on restricted cubic splines and knot placement; @wood2017generalized on regression splines and their basis functions.

:::

## 11.4 Penalized splines and generalized additive models {#s4}

Choosing the number of knots is a choice of flexibility, and choosing it by looking at the results is data dredging. A **penalized spline** uses many knots, then restrains wiggliness with a penalty. A small penalty gives a wiggly curve; a large one shrinks it toward a straight line. The **effective degrees of freedom** measure the flexibility actually used. In a **GAM**, the penalty is estimated from the data. Move the penalty and watch the curve and the AIC.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="gL">Penalty (log scale)</label><input type="range" id="gL" min="-4" max="5" step="0.25" value="1"><output id="gLO"></output></div>
  <div class="btns"><button type="button" id="gBest">Choose the penalty by AIC</button></div>
  <svg id="gPlot" viewBox="0 0 520 300" role="img" aria-label="Penalized spline exposure-response curve"></svg>
</div>
<div class="cpanel"><div class="statgrid" id="gStats"></div><div class="now" id="gNow"></div></div>
</div>
```

::: {.callout-warning title="Common pitfall"}

Trying several numbers of knots, or several sets of cut-points, and reporting the one that gives the most striking result. Fix the flexibility in advance, or let a penalty chosen by a criterion such as AIC or cross-validation decide.

:::

::: {.callout-tip icon="false" title="Bayesian preview"}

A penalty that shrinks spline coefficients toward a straight line is mathematically the same as a prior that says "the curve is probably smooth". In a Bayesian model the penalty becomes a random-walk or smoothing prior whose strength is learned from the data. Lesson 25 uses this for the temperature curve and the time trend.

:::

::: {.further}

**Further reading:** @eilers1996flexible introduced P-splines; @wood2017generalized is the standard book on GAMs and penalty selection; @perperoglou2019review on penalized splines in R.

:::

## Check your understanding {#quiz}

```{=html}
<div class="cw" id="quizzes"></div>
```

## Key takeaways {#take}

- Categorizing a continuous exposure gives a step function that depends on arbitrary cut-points and wastes information.
- Polynomials are global and unstable at the extremes of the data.
- Regression splines are flexible locally and well behaved; natural (restricted) cubic splines are linear beyond the outer knots.
- Penalized splines let the data choose the flexibility, measured by effective degrees of freedom.
- Present nonlinear effects as rate ratios relative to a stated reference, with confidence bands.

```{=html}
<div class="done" data-lesson="lesson-11"><button type="button"></button><span></span></div>
<script src="assets/lesson-11.js"></script>
```
