---
title: "Lesson 13 · Measurement error and misclassification"
subtitle: "Part III · Causal inference, measurement and disease burden"
description: "Classical and Berkson exposure error, regression dilution, validation studies and regression calibration, SIMEX, and misclassification of binary variables."
---

```{r}
#| include: false
source("R/course-data.R")
```

::: {.meta-pills}

[About 3 hours]{} [Before this: Lessons 8 and 12]{} [Threads: bias · uncertainty · frequentist ↔ Bayesian]{}

:::

::: {.objectives}

By the end of this lesson you will be able to

- distinguish classical from Berkson error, and predict the bias each causes;
- explain regression dilution and calculate the attenuation factor;
- use a validation subsample to correct an estimate by regression calibration;
- explain how SIMEX corrects for error by adding more of it;
- predict the effect of non-differential and differential misclassification, and correct for it with assumed sensitivity and specificity.

:::

::: {.why}

**Why it matters.** No air pollution study measures each child's true exposure. Land-use regression models, monitors and questionnaires all err. In our cohort, the true effect of NO₂ is an odds ratio of 1.15 per 10 µg/m³, but the fully adjusted estimate using modeled NO₂ is only about 1.05. Measurement error, not confounding, causes most of that gap. Knowing which kind of error you have tells you whether, and how, to correct.

:::

## 13.1 Classical and Berkson error {#s1}

**Classical error** adds noise to the true value: a noisy monitor, a crude exposure model. It flattens exposure–response slopes toward zero, which is called **regression dilution**. **Berkson error** goes the other way: everyone is assigned a group value, such as their district's average, and the true values scatter around it. It barely biases a linear slope, but it widens the confidence interval. Change the error type and size, and compare the estimate with the truth.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="meT">Exposure used in the analysis</label><select id="meT"><option value="cl">true NO₂ plus classical error</option><option value="bk">district average of true NO₂ (Berkson)</option><option value="mod">the cohort's modeled NO₂</option></select></div>
  <div class="ctrl" id="meRow"><label for="meS">Error standard deviation (µg/m³)</label><input type="range" id="meS" min="0" max="15" step="0.5" value="5"><output id="meSO"></output></div>
  <svg id="mePlot" viewBox="0 0 520 260" role="img" aria-label="Estimated odds ratio against error size"></svg>
  <div class="legend"><span><i style="background:var(--p1)"></i>expected under classical error</span><span><i style="background:var(--p2)"></i>this analysis</span><span><i style="background:var(--p4)"></i>truth</span></div>
</div>
<div class="cpanel"><div class="statgrid" id="meStats"></div><div class="now" id="meNow"></div></div>
</div>
```

::::: {.triad}

:::: {.a}

::: {.lab}

[A]{.k}Standard notation

:::

$$\text{classical: } W = X + U, \quad U \perp X$$
$$\text{Berkson: } X = W + U, \quad U \perp W$$
$$\hat\beta_W \approx \lambda\,\beta, \qquad \lambda = \frac{\sigma^2_{X \mid Z}}{\sigma^2_{X \mid Z} + \sigma^2_U}$$

::: {.note}

$X$ = true exposure, $W$ = measured, $Z$ = covariates; $\lambda$ is the reliability ratio (attenuation factor).

:::

::::

:::: {.b}

::: {.lab}

[B]{.k}Model list and R

:::

```r
# classical error attenuates
glm(asthma ~ I(no2_modeled / 10) + ses + green,
    family = binomial, data = cohort)
# true exposure (simulation only)
glm(asthma ~ I(no2_true / 10) + ses + green,
    family = binomial, data = cohort)
```

::::

:::: {.c}

::: {.lab}

[C]{.k}How it's reported

:::

[ ]{#meReport}

::::

:::::

::: {.further}

**Further reading:** @carroll2006measurement is the standard text; @armstrong1998effect explains classical and Berkson error for environmental exposures; @zeger2000exposure apply these ideas to air pollution studies.

:::

## 13.2 Validation data and regression calibration {#s2}

To correct for classical error you need to know how big it is. A **validation study** measures the exposure better in a subsample: here, 280 children wore personal NO₂ monitors. **Regression calibration** uses the validation data to predict each child's true exposure from the error-prone one, then fits the outcome model with the predicted exposure. Its key ingredient is the calibration slope, which is usually below 1.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="rcN">Children in the validation study</label><input type="range" id="rcN" min="20" max="280" step="10" value="280"><output id="rcNO"></output></div>
  <div class="btns"><button type="button" id="rcNew">Draw a different validation subsample</button></div>
  <svg id="rcPlot" viewBox="0 0 520 280" role="img" aria-label="Personal against modeled NO2 in the validation study"></svg>
</div>
<div class="cpanel"><div class="statgrid" id="rcStats"></div><div class="now" id="rcNow"></div></div>
</div>
```

::::: {.triad}

:::: {.a}

::: {.lab}

[A]{.k}Standard notation

:::

$$E[X \mid W, Z] = \alpha_0 + \alpha_1 W + \alpha_2^\top Z \quad \text{(validation data)}$$
$$\mathrm{logit}\,P(Y = 1) = \beta_0 + \beta\,\widehat{E}[X \mid W, Z] + \gamma^\top Z$$

::::

:::: {.b}

::: {.lab}

[B]{.k}Model list and R

:::

```r
val <- cohort |> filter(validation)
cal <- lm(no2_personal ~ no2_modeled + ses + green,
          data = val)
cohort$no2_cal <- predict(cal, newdata = cohort)
glm(asthma ~ I(no2_cal / 10) + ses + green,
    family = binomial, data = cohort)
```

::::

:::: {.c}

::: {.lab}

[C]{.k}How it's reported

:::

[ ]{#rcReport}

::::

:::::

::: {.callout-note collapse="true" appearance="simple" icon="false" title="R code (optional): regression calibration with a bootstrap interval"}

```{r}
rc_or <- function(d) {
  cal <- lm(no2_personal ~ no2_modeled + ses + green, data = filter(d, validation))
  d$no2_cal <- predict(cal, newdata = d)
  exp(coef(glm(asthma ~ I(no2_cal / 10) + ses + green, family = binomial, data = d))[2])
}
naive <- exp(coef(glm(asthma ~ I(no2_modeled / 10) + ses + green, family = binomial, data = cohort))[2])
c(naive = unname(naive), calibrated = unname(rc_or(cohort)))

set.seed(13)
boot <- map_dbl(1:200, \(b) rc_or(slice_sample(cohort, prop = 1, replace = TRUE)))
quantile(boot, c(0.025, 0.975))
```

:::

::: {.further}

**Further reading:** @rosner1990correction developed regression calibration for logistic regression; @keogh2020stratos give practical guidance on validation studies and simple corrections; @shaw2020stratos cover more complex methods.

:::

## 13.3 SIMEX: correcting by adding more error {#s3}

If the error variance is known, **simulation extrapolation** (SIMEX) estimates what the result would be with no error by first making the error *worse*. Add extra noise of increasing size, refit, and watch the estimate fall. Then extrapolate the trend back to the point where the total error would be zero.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="sxS">Assumed error SD of modeled NO₂</label><input type="range" id="sxS" min="1" max="10" step="0.5" value="5"><output id="sxSO"></output></div>
  <div class="btns"><button type="button" id="sxRun">Run SIMEX</button></div>
  <svg id="sxPlot" viewBox="0 0 520 280" role="img" aria-label="SIMEX extrapolation"></svg>
</div>
<div class="cpanel"><div class="statgrid" id="sxStats"></div><div class="now" id="sxNow"></div></div>
</div>
```

::: {.further}

**Further reading:** @cook1994simulation introduced SIMEX; @carroll2006measurement explain its extrapolation step; @shaw2020stratos compare SIMEX with other corrections.

:::

## 13.4 Misclassification of binary variables {#s4}

For binary variables, measurement error is **misclassification**, described by sensitivity and specificity. Suppose the analysis used the wheeze questionnaire instead of a clinical diagnosis of asthma. If the questionnaire errs equally in high- and low-NO₂ children (**non-differential**), the odds ratio is usually biased toward 1. If parents of highly exposed children report wheeze more readily (**differential**), the bias can go either way. Knowing or assuming sensitivity and specificity, you can correct the counts.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="mcSe">Sensitivity, low-NO₂ children</label><input type="range" id="mcSe" min="50" max="100" value="85"><output id="mcSeO"></output></div>
  <div class="ctrl"><label for="mcSe1">Sensitivity, high-NO₂ children</label><input type="range" id="mcSe1" min="50" max="100" value="85"><output id="mcSe1O"></output></div>
  <div class="ctrl"><label for="mcSp">Specificity (both groups)</label><input type="range" id="mcSp" min="70" max="100" value="90"><output id="mcSpO"></output></div>
  <div class="btns"><button type="button" id="mcND">Make it non-differential</button></div>
  <div class="tw"><table data-quarto-disable-processing="true" class="t" id="mcTab"></table></div>
</div>
<div class="cpanel"><div class="statgrid" id="mcStats"></div><div class="now" id="mcNow"></div></div>
</div>
```

::::: {.triad}

:::: {.a}

::: {.lab}

[A]{.k}Standard notation

:::

$$E[a^*] = Se\cdot a + (1 - Sp)(n - a)$$
$$\hat a = \frac{a^* - (1 - Sp)\, n}{Se + Sp - 1}$$

::: {.note}

$a^*$ = observed positives in a group of size $n$; $\hat a$ = corrected number of true cases.

:::

::::

:::: {.b}

::: {.lab}

[B]{.k}Model list and R

:::

```r
correct <- function(pos, n, se, sp)
  (pos - (1 - sp) * n) / (se + sp - 1)
# or a probabilistic bias analysis:
# episensr::probsens(...)
```

::::

:::: {.c}

::: {.lab}

[C]{.k}How it's reported

:::

[ ]{#mcReport}

::::

:::::

::: {.callout-warning title="Common pitfall"}

Assuming that non-differential error always biases toward the null, so a non-null result must be "conservative". That holds only on average and for binary exposures with independent errors. For exposures with three or more categories, for errors that are correlated with errors in confounders, or in any single study by chance, the estimate can be pushed away from the null.

:::

::: {.callout-tip icon="false" title="Bayesian preview"}

A Bayesian measurement-error model treats each child's true exposure as an unknown parameter, informed by the modeled value, the validation data and its relation to the outcome, all at once. Uncertainty about the error is carried into the final interval automatically. Lesson 24 builds this as a hierarchical model.

:::

::: {.further}

**Further reading:** @jurek2005proper on why non-differential misclassification need not bias a single study toward the null; @fox2021applying on simple and probabilistic bias analysis for misclassification; @keogh2020stratos on misclassification in observational epidemiology.

:::

## Check your understanding {#quiz}

```{=html}
<div class="cw" id="quizzes"></div>
```

## Key takeaways {#take}

- Classical error (noise added to the truth) attenuates slopes toward zero; Berkson error (truth scattered around an assigned value) mostly costs precision.
- The attenuation factor is the share of exposure variance, after accounting for covariates, that is true signal.
- Validation data make correction possible: regression calibration replaces the measured exposure by its expected true value.
- SIMEX adds error deliberately and extrapolates back to zero error.
- Non-differential misclassification usually, not always, biases toward the null; differential misclassification can bias in either direction.

```{=html}
<div class="done" data-lesson="lesson-13"><button type="button"></button><span></span></div>
<script src="assets/lesson-13.js"></script>
```
