---
title: "Lesson 5 · Likelihood and estimation"
subtitle: "Part I · Foundations"
description: "The likelihood function, maximum likelihood estimates, curvature and standard errors, and likelihood ratios."
---

```{r}
#| include: false
source("R/course-data.R")
```

::: {.meta-pills}

[About 2.5 hours]{} [Before this: Lessons 3–4]{} [Threads: likelihood · uncertainty · frequentist ↔ Bayesian]{}

:::

::: {.objectives}

By the end of this lesson you will be able to

- write the likelihood for a risk and for a Poisson rate, and explain what it measures;
- find a maximum likelihood estimate from the shape of the likelihood;
- explain how the curvature of the log-likelihood gives the standard error;
- compare a Wald interval with a likelihood-based interval, and say when they differ;
- use a likelihood ratio to compare how well two parameter values are supported by the data.

:::

::: {.why}

**Why it matters.** Almost every estimate in epidemiology, from an odds ratio out of logistic regression to the heat effect in a DLNM, is a maximum likelihood estimate. The likelihood is also the ingredient Bayesian analysis shares with frequentist analysis. Understanding it on a simple risk makes every later model less of a black box.

:::

## 5.1 The likelihood of a risk {#s1}

Suppose a study follows n children and observes d cases of asthma. For any candidate value of the risk π, the binomial distribution says how probable those data would be. Viewed as a function of π, with the data fixed, that probability is the **likelihood**. Values of π that make the observed data more probable are better supported. The peak is the **maximum likelihood estimate** (MLE).

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="lN">Children followed, n</label><input type="range" id="lN" min="10" max="2000" value="100"><output id="lNO"></output></div>
  <div class="ctrl"><label for="lPi">Candidate risk, π</label><input type="range" id="lPi" min="1" max="40" value="20"><output id="lPiO"></output></div>
  <div class="btns"><button type="button" id="lNew">Draw a new sample</button><label class="small"><input type="checkbox" id="lLog"> show the log-likelihood</label></div>
  <svg id="lik" viewBox="0 0 520 270" role="img" aria-label="Likelihood of the risk"></svg>
</div>
<div class="cpanel"><div class="statgrid" id="lStats"></div><div class="now" id="lNow"></div></div>
</div>
```

::::: {.triad}

:::: {.a}

::: {.lab}

[A]{.k}Standard notation

:::

$$L(\pi) = \binom{n}{d}\pi^{d}(1-\pi)^{n-d}$$
$$\ell(\pi) = d\log\pi + (n-d)\log(1-\pi) + \text{const}$$
$$\hat\pi = \frac{d}{n}$$

::::

:::: {.b}

::: {.lab}

[B]{.k}Model list and R

:::

```r
d ~ binomial(n, pi)

loglik <- function(p) dbinom(d, n, p, log = TRUE)
optimize(loglik, c(0, 1), maximum = TRUE)
```

::::

:::: {.c}

::: {.lab}

[C]{.k}How it's reported

:::

[ ]{#lReport}

::::

:::::

::: {.callout-warning title="Common pitfall"}

Reading the likelihood as the probability that π takes a given value. L(π) is the probability of the *data* given π. It doesn't sum to 1 over π, and turning it into a probability for π requires a prior: that is what Bayesian inference does.

:::

::: {.further}

**Further reading:** @pawitan2001all is a thorough modern treatment of likelihood; @clayton1993statistical build epidemiological methods on it; @held2020applied develop likelihood and Bayesian inference side by side.

:::

## 5.2 The likelihood of a Poisson rate {#s2}

For count data, each day's count has a Poisson probability given the rate λ, and if days are independent the likelihood is the product of these probabilities over all days. In practice we add their logarithms. Choose a district, a number of days, and a candidate rate: each bar below is one day's probability under that rate, and the total log-likelihood combines them.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="pD">District (age 65+)</label><select id="pD"></select></div>
  <div class="ctrl"><label for="pN">Days observed</label><input type="range" id="pN" min="5" max="120" value="30"><output id="pNO"></output></div>
  <div class="ctrl"><label for="pL">Candidate rate λ (visits/day)</label><input type="range" id="pL" min="1" max="60" step="0.1" value="10"><output id="pLO"></output></div>
  <svg id="pBars" viewBox="0 0 520 200" role="img" aria-label="Each day's probability under the candidate rate"></svg>
  <svg id="pLik" viewBox="0 0 520 220" role="img" aria-label="Log-likelihood of the rate"></svg>
</div>
<div class="cpanel"><div class="statgrid" id="pStats"></div><div class="now" id="pNow"></div></div>
</div>
```

::::: {.triad}

:::: {.a}

::: {.lab}

[A]{.k}Standard notation

:::

$$y_t \sim \mathrm{Poisson}(\lambda), \quad t = 1, \dots, n$$
$$\ell(\lambda) = \sum_{t} \left[ y_t \log\lambda - \lambda \right] + \text{const}, \qquad \hat\lambda = \bar y$$

::::

:::: {.b}

::: {.lab}

[B]{.k}Model list and R

:::

```r
visits_t ~ poisson(lambda)

ll <- function(l) sum(dpois(y, l, log = TRUE))
glm(y ~ 1, family = poisson) |>
  coef() |> exp()          # equals mean(y)
```

::::

:::: {.c}

::: {.lab}

[C]{.k}How it's reported

:::

[ ]{#pReport}

::::

:::::

::: {.callout-note collapse="true" appearance="simple" icon="false" title="R code (optional): a Poisson likelihood by hand and with glm()"}

```{r}
y <- visits |>
  filter(district == "Old Town", age_group == "65+") |>
  slice_head(n = 30) |>
  pull(visits)

ll <- function(l) sum(dpois(y, l, log = TRUE))
grid <- tibble(lambda = seq(15, 35, by = 0.1)) |>
  mutate(loglik = map_dbl(lambda, ll))
grid |> slice_max(loglik)            # the peak of the curve

fit <- glm(y ~ 1, family = poisson)
exp(coef(fit))                        # the same MLE: mean(y)
```

:::

::: {.further}

**Further reading:** @clayton1993statistical on the Poisson likelihood for rates; @pawitan2001all on likelihoods for independent observations; @cameron2013regression for Poisson models of counts.

:::

## 5.3 Curvature, standard errors and likelihood intervals {#s3}

How sharply the log-likelihood peaks tells you how precisely the data pin down the parameter. The curvature at the peak is the **information**; its inverse square root is the **standard error**. Two kinds of 95% interval follow. The **Wald interval** is the MLE ± 1.96 SE: symmetric by construction. The **likelihood interval** contains all values within 1.92 log-likelihood units of the peak. With few cases and a small risk, the likelihood is lopsided, and the two disagree.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="cD">Cases, d</label><input type="range" id="cD" min="0" max="40" value="3"><output id="cDO"></output></div>
  <div class="ctrl"><label for="cN">Children, n</label><input type="range" id="cN" min="20" max="400" value="40"><output id="cNO"></output></div>
  <svg id="curv" viewBox="0 0 520 270" role="img" aria-label="Log-likelihood with its quadratic approximation"></svg>
  <div class="legend"><span><i style="background:var(--c-lik)"></i>log-likelihood</span><span><i style="background:var(--p1)"></i>quadratic approximation (Wald)</span><span><i style="background:var(--p2)"></i>1.92 below the peak</span></div>
</div>
<div class="cpanel"><div class="statgrid" id="cStats"></div><div class="now" id="cNow"></div></div>
</div>
```

::::: {.triad}

:::: {.a}

::: {.lab}

[A]{.k}Standard notation

:::

$$I(\hat\pi) = -\ell''(\hat\pi) = \frac{n}{\hat\pi(1-\hat\pi)}, \qquad SE = \frac{1}{\sqrt{I(\hat\pi)}}$$
$$\text{Wald: } \hat\pi \pm 1.96\,SE \qquad \text{Likelihood: } \{\pi : \ell(\hat\pi) - \ell(\pi) \le 1.92\}$$

::::

:::: {.b}

::: {.lab}

[B]{.k}Model list and R

:::

```r
se <- sqrt(p_hat * (1 - p_hat) / n)
p_hat + c(-1.96, 1.96) * se     # Wald

fit <- glm(cbind(d, n - d) ~ 1,
           family = binomial)
confint(fit) |> plogis()         # profile likelihood
```

::::

:::: {.c}

::: {.lab}

[C]{.k}How it's reported

:::

[ ]{#cReport}

::::

:::::

::: {.further}

**Further reading:** @pawitan2001all on information and likelihood-based intervals; @held2020applied compare Wald, likelihood and other intervals; @clayton1993statistical show the quadratic approximation for epidemiological parameters.

:::

## 5.4 Likelihood ratios: comparing two values {#s4}

The likelihood ratio L(π₁)/L(π₂) says how many times better the data support π₁ than π₂. A ratio of 8 is often called fairly strong support, and 32 strong support. It depends only on the data actually observed, not on data that might have been, and it is the basis of the likelihood ratio test (Lesson 10) and of Bayes factors (Lesson 23).

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="rD">Cases observed</label><input type="range" id="rD" min="0" max="60" value="26"><output id="rDO"></output></div>
  <div class="ctrl"><label for="rN">Children followed</label><input type="range" id="rN" min="50" max="1000" value="200"><output id="rNO"></output></div>
  <div class="ctrl"><label for="rA">Hypothesis 1: risk</label><input type="range" id="rA" min="1" max="40" value="10"><output id="rAO"></output></div>
  <div class="ctrl"><label for="rB">Hypothesis 2: risk</label><input type="range" id="rB" min="1" max="40" value="15"><output id="rBO"></output></div>
  <svg id="lr" viewBox="0 0 520 240" role="img" aria-label="Likelihood with two hypotheses marked"></svg>
</div>
<div class="cpanel"><div class="statgrid" id="rStats"></div><div class="now" id="rNow"></div></div>
</div>
```

::: {.callout-tip icon="false" title="Bayesian preview"}

Multiply the likelihood curve by a prior distribution for π, rescale so the area is 1, and you have the posterior distribution. With a flat prior, the posterior has exactly the shape of the likelihood, and its peak is the MLE. The likelihood is the part of the analysis on which frequentists and Bayesians agree. Lesson 19 builds on this.

:::

::: {.further}

**Further reading:** @royall1997statistical develops the likelihood ratio as a measure of evidence and proposes the benchmarks of 8 and 32; @pawitan2001all and @held2020applied relate likelihood ratios to tests and Bayes factors.

:::

## Check your understanding {#quiz}

```{=html}
<div class="cw" id="quizzes"></div>
```

## Key takeaways {#take}

- The likelihood is the probability of the observed data, viewed as a function of the parameter.
- The maximum likelihood estimate is the parameter value that makes the data most probable. For a risk it is d/n; for a Poisson rate it is the mean count.
- The curvature of the log-likelihood at its peak gives the standard error: sharper peak, smaller SE.
- Wald intervals assume a symmetric, quadratic log-likelihood. With few events, likelihood-based intervals are more accurate.
- Likelihood ratios compare how well the data support two parameter values.

```{=html}
<div class="done" data-lesson="lesson-05"><button type="button"></button><span></span></div>
<script src="assets/lesson-05.js"></script>
```
