---
title: "Lesson 10 · Model checking, comparison and selection"
subtitle: "Part II · Frequentist regression"
description: "Building and checking a model step by step, likelihood ratio tests and AIC, overfitting and cross-validation, influence, and why stepwise selection misleads."
---

```{r}
#| include: false
source("R/course-data.R")
```

::: {.meta-pills}

[About 2.5 hours]{} [Before this: Lessons 6 and 8]{} [Threads: uncertainty · frequentist ↔ Bayesian]{}

:::

::: {.objectives}

By the end of this lesson you will be able to

- use residuals, autocorrelation and dispersion to judge whether a model is adequate;
- compare nested models with a likelihood ratio test, and any models with AIC or QAIC;
- explain overfitting and estimate out-of-sample error with cross-validation;
- identify influential observations with leverage and Cook's distance;
- explain why data-driven variable selection produces misleading p-values.

:::

::: {.why}

**Why it matters.** A model can fit the data you have while describing the world badly. Leftover seasonal patterns can masquerade as temperature effects; a single extreme day can move a slope; and choosing variables by their p-values guarantees that some "findings" are noise. Checking and comparing models carefully is what separates an analysis from a fishing trip.

:::

## 10.1 Building and checking a model step by step {#s1}

Here is the daily count of emergency visits in the whole city, modeled with Poisson regression. Add terms one at a time and watch three diagnostics: the **residuals over time** (patterns mean something is missing), their **autocorrelation** (today's residual predicting tomorrow's), and the **dispersion**. The table compares each term's contribution with a likelihood ratio test.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <p class="small">Terms in the model of daily city-wide visits:</p>
  <label class="small"><input type="checkbox" id="tDow"> day of week</label>
  <label class="small"><input type="checkbox" id="tSea"> season (annual cycle)</label>
  <label class="small"><input type="checkbox" id="tTrend"> long-term trend</label>
  <label class="small"><input type="checkbox" id="tTemp"> temperature (spline)</label>
  <label class="small"><input type="checkbox" id="tPM"> PM2.5</label>
  <svg id="tRes" viewBox="0 0 520 200" role="img" aria-label="Residuals over time"></svg>
  <svg id="tAcf" viewBox="0 0 520 150" role="img" aria-label="Autocorrelation of residuals"></svg>
</div>
<div class="cpanel"><div class="statgrid" id="tStats"></div><div class="tw"><table data-quarto-disable-processing="true" class="t" id="tTab"></table></div><div class="now" id="tNow"></div></div>
</div>
```

::::: {.triad}

:::: {.a}

::: {.lab}

[A]{.k}Standard notation

:::

$$LR = 2\,(\ell_1 - \ell_0) \sim \chi^2_{k} \ \text{under } H_0$$
$$AIC = -2\ell + 2p, \qquad QAIC = -\frac{2\ell}{\hat\phi} + 2p$$

::: {.note}

$k$ = parameters added; $p$ = parameters in the model. For quasi-Poisson models, the LR statistic is divided by $\hat\phi$.

:::

::::

:::: {.b}

::: {.lab}

[B]{.k}Model list and R

:::

```r
m0 <- glm(visits ~ dow, family = quasipoisson,
          data = city_daily)
m1 <- update(m0, . ~ . + sin(2*pi*doy/365.25) +
               cos(2*pi*doy/365.25))
anova(m0, m1, test = "F")   # quasi-Poisson
acf(residuals(m1, type = "pearson"))
```

::::

:::: {.c}

::: {.lab}

[C]{.k}How it's reported

:::

[ ]{#tReport}

::::

:::::

::: {.callout-note collapse="true" appearance="simple" icon="false" title="R code (optional): building the model and checking residual autocorrelation"}

```{r}
city_daily <- visits |>
  group_by(date) |>
  summarise(visits = sum(visits), temp = first(temp_city), pm25 = first(pm25),
            dow = first(dow), .groups = "drop") |>
  mutate(doy = yday(date), t = as.numeric(date - min(date)))

m_dow  <- glm(visits ~ dow, family = quasipoisson, data = city_daily)
m_full <- glm(visits ~ dow + sin(2*pi*doy/365.25) + cos(2*pi*doy/365.25) + t +
                splines::ns(temp, df = 4) + pm25, family = quasipoisson, data = city_daily)
anova(m_dow, m_full, test = "F")
summary(m_full)$dispersion
acf(residuals(m_full, type = "pearson"), lag.max = 21, plot = FALSE)$acf[2:8]
```

:::

::: {.further}

**Further reading:** @harrell2015regression on checking regression models; @fox2016applied on GLM diagnostics; @mccullagh1989generalized on deviance and likelihood ratio tests.

:::

## 10.2 Overfitting and cross-validation {#s2}

A more flexible model always fits the data it was estimated on better. The question is whether it predicts **new** data better. Here a polynomial in temperature is fitted to a small training sample of days, and its predictions are scored on all the other days. Increase the degree: the training error keeps falling, the test error eventually rises. That gap is **overfitting**. **Cross-validation** estimates the test error using only the training data.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="oD">Polynomial degree</label><input type="range" id="oD" min="1" max="10" value="2"><output id="oDO"></output></div>
  <div class="ctrl"><label for="oN">Training days</label><input type="range" id="oN" min="15" max="400" value="30"><output id="oNO"></output></div>
  <div class="btns"><button type="button" id="oNew">New training sample</button></div>
  <svg id="oFit" viewBox="0 0 520 240" role="img" aria-label="Polynomial fit to training days"></svg>
  <svg id="oErr" viewBox="0 0 520 200" role="img" aria-label="Training, cross-validation and test error by degree"></svg>
  <div class="legend"><span><i style="background:var(--p1)"></i>training error</span><span><i style="background:var(--p4)"></i>5-fold cross-validation</span><span><i style="background:var(--p2)"></i>error on new days</span></div>
</div>
<div class="cpanel"><div class="statgrid" id="oStats"></div><div class="now" id="oNow"></div></div>
</div>
```

::: {.further}

**Further reading:** @hastie2009elements on the bias–variance trade-off and cross-validation; @harrell2015regression on overfitting and validation; @burnham2002model on information criteria as estimates of predictive performance.

:::

## 10.3 Influential observations {#s3}

Least squares squares the residuals, so a single unusual point can pull a line strongly. Influence combines an unusual **outcome** (a large residual) with an unusual **predictor value** (high **leverage**). Move the extra child around and watch the slope and **Cook's distance**, which measures how much the fitted values change when that point is removed.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="iX">Extra child's green space</label><input type="range" id="iX" min="0" max="1.6" step="0.01" value="0.5"><output id="iXO"></output></div>
  <div class="ctrl"><label for="iY">Extra child's NO₂</label><input type="range" id="iY" min="0" max="90" step="0.5" value="30"><output id="iYO"></output></div>
  <svg id="iPlot" viewBox="0 0 520 280" role="img" aria-label="Regression with one movable point"></svg>
  <div class="legend"><span><i style="background:var(--c-lik)"></i>line without the extra child</span><span><i style="background:var(--p2)"></i>line with the extra child</span></div>
</div>
<div class="cpanel"><div class="statgrid" id="iStats"></div><div class="now" id="iNow"></div></div>
</div>
```

::: {.further}

**Further reading:** @fox2016applied on leverage, residuals and Cook's distance; @harrell2015regression on influential observations; @kirkwood2003essential on outliers in medical data.

:::

## 10.4 Why stepwise selection misleads {#s4}

Choosing variables because they are "significant" in the same data used to report them makes those p-values meaningless. Here, 20 candidate predictors of asthma are pure noise. A screening rule keeps those with p < 0.05. Run it many times: noise variables are selected regularly, and every selected one comes with a "significant" p-value, by construction.

```{=html}
<div class="cw grid2">
<div class="cpanel">
  <div class="ctrl"><label for="vK">Noise candidates</label><input type="range" id="vK" min="5" max="50" value="20"><output id="vKO"></output></div>
  <div class="ctrl"><label for="vN">Children per study</label><input type="range" id="vN" min="100" max="2000" step="50" value="300"><output id="vNO"></output></div>
  <div class="btns"><button type="button" id="vRun">Run 500 studies</button></div>
  <svg id="vPlot" viewBox="0 0 520 220" role="img" aria-label="Number of noise variables selected"></svg>
</div>
<div class="cpanel"><div class="statgrid" id="vStats"></div><div class="now" id="vNow"></div></div>
</div>
```

::: {.callout-warning title="Common pitfall"}

Selecting confounders by stepwise regression or by "a change in estimate of more than 10%", then reporting the final model's confidence intervals as if it had been chosen in advance. Confounders should be chosen from causal knowledge (a DAG, Lesson 9), and the model fixed before looking at the outcome.

:::

::: {.callout-tip icon="false" title="Bayesian preview"}

Bayesian analysis compares models in two ways: by Bayes factors, which ask which model predicted the data better before seeing it, and by leave-one-out cross-validation, the Bayesian counterpart of AIC. Lesson 23 compares them, and shows why priors, not stepwise rules, are the natural way to handle many candidate predictors.

:::

::: {.further}

**Further reading:** @heinze2018variable review variable selection and its pitfalls; @greenland1989modeling on modeling strategy in epidemiology; @harrell2015regression on the problems of stepwise selection.

:::

## Check your understanding {#quiz}

```{=html}
<div class="cw" id="quizzes"></div>
```

## Key takeaways {#take}

- Residual patterns over time, residual autocorrelation and overdispersion show what a model is missing.
- Nested models are compared with likelihood ratio tests; any models fitted to the same data can be compared with AIC (QAIC for quasi-Poisson). Lower is better.
- More complex models always fit the training data better. Cross-validation estimates how well they predict new data.
- Influential points combine high leverage with large residuals. Check them; don't silently delete them.
- Data-driven selection, such as stepwise regression, invalidates the p-values and intervals of the final model.

```{=html}
<div class="done" data-lesson="lesson-10"><button type="button"></button><span></span></div>
<script src="assets/lesson-10.js"></script>
```
